---
title: "ME - GIA. PCA Introduction (2)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    toc: true
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In the 1988 Olympics held in Seoul, the heptathlon was won by one of the stars 
of woman's athletics in the USA, Jackie Joyner-Kersee.

The heptathlon data set in the R package HSAUR3 contains the results for all 25 
competitors in all seven disciplines.

We will apply PCA with a view to exploring the structure of these data and assessing 
how the PCs relate to the scores assigned by the official scoring system.

First, we load the packages we will need. 

```{r message=FALSE,warning=FALSE}
library(dplyr)
library(kableExtra)
library(HSAUR3)
library(ggplot2)
library(GGally)
library(factoextra)

# Load data
data(heptathlon)
```

Let's print the table.

```{r, include=TRUE}
heptathlon %>%
  kbl(caption = "Heptathlon results in 1988 Olympics held in Seoul") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```


You can observe that the scores of the different events do not have the same direction: 

- The events *hurdles, run200m, and run800m*, **the lower value the better**. 
- The other events have a different direction, **the more, the better**.

Thus, we change the direction of the *hurdles, run200m, and run800m* to have all 
seven events in the same direction

```{r, include=TRUE}
heptathlon$hurdles <- with(heptathlon, max(hurdles) - hurdles)
heptathlon$run200m <- with(heptathlon, max(run200m) - run200m)
heptathlon$run800m <- with(heptathlon, max(run800m) - run800m)
```

Now we have all events in the same direction and all are numeric variables. Therefore, we can graph scatter plots in pairs. 

```{r, include=TRUE}
score <- which(colnames(heptathlon) == "score")
plot(heptathlon[, -score])
```

The correlation matrix (all attributes are numeric)

```{r, include=TRUE}
round(cor(heptathlon[,-score]), 2) %>% 
  kbl() %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

```{r, include=TRUE, warning=FALSE}
ggpairs(heptathlon[, -score], progress = FALSE)
```

Laura (PNG) (last row) is the outlier (values 0 in hurdles and run800m). Check also the left-skewed tails in the density plots. We remove it.

```{r, include=TRUE}
heptathlon <- heptathlon[-grep("PNG", rownames(heptathlon)), ]
```

And now check the scatter plots and correlation matrix again:
```{r, include=TRUE, warning=FALSE}
ggpairs(heptathlon[, -score], progress = FALSE)
```

## PCA using princomp

Thus, we run the PCA using princomp and stating that we will use the correlation matrix. 

```{r, include=TRUE}
hept_def       <- scale(heptathlon[, -score])
heptathlon_pca <- princomp(hept_def)
```


Let's check the results step by step. The coefficients (also called loadings or loading vector) for the first PC are obtained as:

```{r, include=TRUE}
a1 <- heptathlon_pca$loadings[,1]
a1
```


#### The variance explained by the principal components

The summary function reports the standard deviation, proportion of
variance, and the cumulative proportion of variance

```{r, include=TRUE}
summary(heptathlon_pca)
```

#### Criteria for choosing the number of components

We plot the *scree plot*.

```{r, include=TRUE}
screeplot(heptathlon_pca, type="l", main="Screeplot heptathlon")
```

#### Principal component scores

The PC scores are the values of each PCs (linear combination of the original variables) evaluated in the data set.

In this example, we can use the PCA scores to see if can predict well the total number of points in the heptathlon competition.

```{r, include=TRUE}
head(heptathlon_pca$scores) %>%
  kbl(caption = "Heptathlon results in 1988 Olympics held in Seoul") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

We can check that the PCs are uncorrelated (all values are 0)

```{r, include=TRUE}
cor(heptathlon_pca$scores) %>%
  kbl(caption = "Heptathlon results in 1988 Olympics held in Seoul") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

Let's finally graph if the first PC1 scores can predict the Olympic heptathlon total number of points. 


```{r, include=TRUE}
pc1.scores <- heptathlon_pca$scores[,1]
plot(heptathlon$score, pc1.scores)
```

It predicts very well only with the first PC because PC1 explains quite a lot of the total of the original variance.

#### PCA biplots

```{r, include=TRUE, fig.height=8}
biplot(heptathlon_pca)
```

It does not look very well because there is text overlapping. Let's improve the graph:

```{r message=FALSE,warning=FALSE}
fviz_eig(heptathlon_pca)
fviz_pca_ind(heptathlon_pca,
             col.ind = "contrib", # Color by the contribution 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE,     # Avoid text overlapping
             title="Athletes heptathlon - PCA")
```


Graph of variables (heptathlon events). Positive correlated variables point to the same side of the plot. Negative correlated variables point to opposite sides of the graph.

```{r, include=TRUE}
fviz_pca_var(heptathlon_pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE,     # Avoid text overlapping
             title="Heptathlon events - PCA")
```

A variable (event) plot (a.k.a. **loading plot**) shows how strongly each event influences a principal component.

What is graphed in the loading plot is the rescaled loadings
```{r, include=TRUE}
rbind(a1 * heptathlon_pca$sdev[1], a2 * heptathlon_pca$sdev[2])
```

PCA Biplot of heptathlon athletes and events simultaneously 

```{r, include=TRUE}
fviz_pca_biplot(heptathlon_pca, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969",  # Individuals color
             title="Heptathlon biplot - PCA")
```

A PCA biplot = PCA score plot + loading plot

For example, look at the top left corner: the athletes Choubenkova (URSS), John (GDR), and Behmer (GDR), and the event run800m are depicted together. If we check the data we see that those three athletes stood out in that event. Or look on Joyner-Kersee (USA), who stood out in long jump (the top of all athletes).

```{r, include=TRUE}
heptathlon %>%
  kbl(caption = "Heptathlon results in 1988 Olympics held in Seoul") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

