---
title: "ME - GIA. Binary (Disaggregated vs. Aggregated)"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  html_document:
    code_folding: show
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
---


```{r, warning=FALSE, message=FALSE}
data <- read.csv("base_trabajo.csv")

data$Blood.Pressure <- as.numeric(as.character(data$Blood.Pressure))
data$Insulin.Levels <- as.numeric(as.character(data$Insulin.Levels))
data$Blood.Glucose.Levels <- as.numeric(as.character(data$Blood.Glucose.Levels))
data$BMI <- as.numeric(as.character(data$BMI))
data$Cholesterol.Category <- as.factor(data$Cholesterol.Category)
data$Cholesterol.Category <- relevel(data$Cholesterol.Category, ref = "Low")
data$Cholesterol.Category <- ifelse(data$Cholesterol.Category == "High", 1, 
                                ifelse(data$Cholesterol.Category == "Low", 0, NA))

library(car)



```


```{r}
summary(data)
```


```{r, fig.height=3}
par(mfrow=c(1,2),las=1)
boxplot(Age ~Cholesterol.Category, data=data, col=2:3)
boxplot(Blood.Glucose.Levels~Cholesterol.Category, data=data, col=2:3)
```


```{r, fig.height=6}
par(mfrow=c(2,2),las=1)
plot(Cholesterol.Category~Blood.Pressure,       data=data, col=3:2)
plot(Cholesterol.Category~Insulin.Levels,       data=data, col=3:2)
plot(Cholesterol.Category~Blood.Glucose.Levels,       data=data, col=3:2)
plot(Cholesterol.Category~BMI,       data=data, col=3:2)
plot(Cholesterol.Category~Age,       data=data, col=3:2)



```


# Models

## Disaggregated data

```{r}
#m_Age_dis <- glm(Cholesterol.Category ~ Age, data = data, family = binomial(link= "logit"))
#summary(m_Age_dis)
#residualPlot(m_Age_dis)

m_extended <- glm(Cholesterol.Category ~ Age * Blood.Pressure + Age * BMI + Age * Insulin.Levels, data = data, family = binomial(link = "logit"))
summary(m_extended)
residualPlot(m_extended)

#anova(m_Age_dis, m_extended, test = "Chisq")
# Comparación de AIC
#AIC(m_Age_dis, m_extended)

```

When disaggregated data is used then $m_i = 1$ and asymptotic theory does not 
apply. Deviance and Pearson statistics do not follow a $\chi^2$ distribution
under the $H_0$ of well-fitted model.

As an informal rule of thumb at a fully disaggregated level for data, a model 
could be good if Deviance statistic is similar to the degrees of freedom of the 
model. Large values indicate lack of fit. 

```{r}
c('Deviance'           = m_extended$deviance,
  'Degrees of freedom' = m_extended$df.res)
```

**Does the model fit well the data?**

**Deviance >> Degrees of freedom $\rightarrow$ The model does not seem to fit well the data**.

## Aggregated data

An aggregated dataset is created using classes defined by age. We use the fucntion
`aggregate`.
**In this case, it is possible to aggregate by a continuous variable because it 
is discretized and the sample size is large enough**.

```{r}
dfAge <- with(data, aggregate(x = cbind(ypos = Cholesterol.Category, yneg = 1 - Cholesterol.Category),
                               by = list(Insulin.Levels = Insulin.Levels, Age = Age, BMI = BMI, Blood.Pressure = Blood.Pressure),
                               FUN = sum))

tail(dfAge,4)

```

Fit the model with aggregated data and compare with the model with disaggregated 
data.

```{r}
m_Age_agg_interactions <- glm(cbind(ypos, yneg) ~ Age * Blood.Pressure + Age * BMI + Age * Insulin.Levels, 
                                  data = dfAge, 
                                  family = binomial(link= "logit"))
summary(m_Age_agg_interactions)

# Ver el resumen del modelo para evaluar la significancia de las variables

# Graficar los residuos para evaluar el ajuste del modelo
residualPlot(m_Age_agg_interactions)



```

In the model, the coefficients are the same that those ones with disaggregated 
data, but not the deviance and degrees of freedom.

```{r}
data.frame(
  dis = round(c(coef(m_extended),
                deviance = m_extended$deviance,
                df = m_extended$df.residual), 3),  # Modelo disaggregated (dis)
  agg = round(c(coef(m_Age_agg_interactions),
                deviance = m_Age_agg_interactions$deviance,
                df = m_Age_agg_interactions$df.residual), 3)  # Modelo con interacciones (agg)
)
```



With aggregated data, we can use a formal test using the deviance statistic 
to check if the model fits well the data.

\begin{center}
$\bigg\{ \begin{array}{l} H_0:\text{The current model fits properly the data} \\ H_1: \text{The current model does not fit properly the data} \end{array}$ 
\end{center}

```{r}
deviance <- m_Age_agg_interactions$deviance # deviance
df       <- m_Age_agg_interactions$df.res   # degrees of freedom
1 - pchisq(deviance,df)        # p-value for hypothesis test
```

**Does the model fit well the data?**

**p < 0.05 $\rightarrow$ The model does not seem to fit well the data**.

## Comparison with null model

The difference of deviances is the same in the two models. Thus, the statistic 
$\Delta D$ is appropriate to compare nested models using function `anova`.

```{r}
# Null model with disaggregated data
m_Age_dis0 <- glm(Cholesterol.Category ~ 1,
                  data   = data, 
                  family = binomial(link= "logit"))

# Null model with aggregated data
m_Age_agg0 <- glm(cbind(ypos,yneg) ~ 1, 
                  data   = dfAge,   
                  family = binomial(link= "logit"))

# Comparison with null model
anova(m_Age_dis0, m_extended, test='Chisq')
anova(m_Age_agg0, m_Age_agg_interactions, test='Chisq')
```

