---
title: "ME - GIA. Binary (Separation problem)"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  html_document:
    code_folding: show
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo    = TRUE,  # Show code?
                      warning = FALSE) # Show warnings?
```

```{r, warning=FALSE, message=FALSE}
library(car)
library(logistf)      # Penalized regression

```

```{r}
d <- read.csv("base_trabajo.csv")
d$Cholesterol.Category <- ifelse(d$Cholesterol.Category == "High", 1, 0)


```


# Separation problem

The separation problem occurs in logistic regression when the predictor variables 
can perfectly predict the outcome, causing the model to have infinitely large or 
undefined coefficients. This prevents the model from converging and producing 
reliable estimates.

Check if there is a **quasi-separation** or a **separation** issue.

```{r}
# Discretizar Blood.Glucose.Levels, BMI, y Blood.Pressure en intervalos
d$GlucoseGroup <- cut(d$Blood.Glucose.Levels, breaks = 5) # Agrupa glucosa en 5 intervalos
d$BMIGroup <- cut(d$BMI, breaks = 3)                      # Agrupa BMI en 3 intervalos
d$PressureGroup <- cut(d$Blood.Pressure, breaks = 3)      # Agrupa presión arterial en 3 intervalos

# Crear la tabla con las variables discretizadas
table_discrete <- with(d, table(Cholesterol.Category, GlucoseGroup, BMIGroup, PressureGroup))
print(table_discrete)

```
# Models

## GzLM model

The GzLM using classic ML estimation does not converge (standard errors are too
large).

```{r}
mod_glm <- glm(Cholesterol.Category ~ . - Target - Cholesterol.Levels, 
               data = d, family = binomial(link="logit"))
summary(mod_glm)

```

R returns some warnings that are not printed in the HTML version.

## Penalized regression

Firth model converges.

```{r}
mod_firth <- logistf(Cholesterol.Category ~ . - Target - Cholesterol.Levels, 
                     data = d, pl = FALSE)
options(digits=2)
summary(mod_firth)
options(digits=7)
```

## Model Comparison

We can calculate confidence intervals using penalized regression.

```{r message=FALSE}
confint(mod_glm)
confint(mod_firth, method = "profile")
```

