---
title: "ME - GIA. Binary (Model selection)"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  html_document:
    code_folding: show
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE,     # Show code?
                      fig.width=10,  # Fig.width
                      fig.height=9)  # Fig.height
```



```{r, warning=FALSE, message=FALSE}
# Load packages
library(effects)
library(car)           # function Anova
library(emmeans)       # function emmeans
library(multcomp)      # function cld
library(multcompView)  # function cld    
library(dplyr)         # data management
library(logistf)

```

```{r, warning=FALSE, message=FALSE}
# Read data
base <- read.csv("C:/Users/Pablo/Desktop/Carpetas/Uni/3r cuatri/ME/trabajo/base_trabajo.csv")
base <- base %>%  
  mutate(Cholesterol.Category = as.factor(Cholesterol.Category))
base$Physical.Activity <- as.factor(base$Physical.Activity)
base$Dietary.Habits <- as.factor(base$Dietary.Habits)
base$Ethnicity <- as.factor(base$Ethnicity)
base$Socioeconomic.Factors <- as.factor(base$Socioeconomic.Factors)
base$Smoking.Status <- as.factor(base$Smoking.Status)
base$Alcohol.Consumption <- as.factor(base$Alcohol.Consumption)
base$History.of.PCOS <- as.factor(base$History.of.PCOS)
base$Previous.Gestational.Diabetes <- as.factor(base$Previous.Gestational.Diabetes)
base$Liver.Function.Tests <- as.factor(base$Liver.Function.Tests)
base$Urine.Test <- as.factor(base$Urine.Test)
base$Early.Onset.Symptoms <- as.factor(base$Early.Onset.Symptoms)
base$Target <- as.factor(base$Target)
base$Environmental.Factors <- as.factor(base$Environmental.Factors)
base$Genetic.Markers <- as.factor(base$Genetic.Markers)
base$Autoantibodies <- as.factor(base$Autoantibodies)
base$Family.History <- as.factor(base$Family.History)
base$Cholesterol.Category <- relevel(base$Cholesterol.Category, ref = "Low")
base$Cholesterol.Category <- ifelse(base$Cholesterol.Category == "High", 1, 0)


head(base)


```

# Descriptive analysis

## Univariate exploratory data analysis

```{r}
summary(base)
table(base$Cholesterol.Category)
```

## Bivariate descriptive analysis

```{r}
vars <- colnames(base)[-which(colnames(base) == "Cholesterol.Category")] 
par(mfrow=c(4,5), mar=c(3,3,3,1)) # 4x5 windows with margins 3,3,3,1
colors <- c(2,3)                  # colors: red (2) and green (3)

for (v in vars) {
  tryCatch({
    if (!is.factor(base[[v]])) {
      form <- as.formula(paste0(v, "~Cholesterol.Category"))
      boxplot(form, base, main = v, col = colors, horizontal = TRUE)
    } else {
      tab <- prop.table(table(base$Cholesterol.Category, base[[v]]), 2)
      barplot(tab, main = v, col = colors)
    }
  }, error = function(e) {
    message("Error en la variable: ", v, " - ", e$message)
  })
}


```

# Variable selection (VS)

## VS using Deviance test

Deviance test using `Anova`function.

```{r}
m0 <- glm(Cholesterol.Category ~ .-Target-Cholesterol.Levels, base, family = binomial(link = 'logit'))
summary(m0)

```

**Interpretation of the coefficients**:

- Positive coefficients in a category of a **factor** indicate a higher 
probability of positive response compared to the reference and vice versa.
- Positive coefficients in a **covariate** indicate a higher probability of 
positive response with higher values and vice versa

The function `Anova` compares the full model with the full model without each
variable using the Deviance test (=`LR`), i.e, for assess `sex`variable, the next 
two models are compared:

**M1**: `Y ~  age + race + height + weight + waist + SBP + DBP + siblings +` 
        `     household + income + insurance + arrested + sleep + glasses + fastfood +` 
        `     caffeine + sugar + cigars + TV + Phealth`

**M2**: `Y ~  age + race + height + weight + waist + SBP + DBP + siblings +` 
        `     household + income + insurance + arrested + sleep + glasses + fastfood +` 
        `     caffeine + sugar + cigars + TV + Phealth + sex`          

```{r}
# Marginal contribution of each variable
Anova(m0, test.statistic = 'LR')
```

**Most variables are significant. We could keep in the model all variables that are significant**.
However is a best option to remove variables one by one. 

```{r}
summary(m0.1 <- glm(Cholesterol.Category ~ log(Age + 1) * BMI + log(Blood.Pressure + 1) *
                                 Waist.Circumference + Insulin.Levels + log(Blood.Glucose.Levels + 1) +
                                 Pancreatic.Health + Birth.Weight,
                                 data = base, family = binomial(link = 'logit')))

# Marginal contribution of each variable
Anova(m0.1, test.statistic = 'LR')
```

## VS using AIC/BIC

We can perform automatic variable selection using the `step` function and **BIC** 
criterion. We will compare with the previous model.

***We can use automatic selection because there are no missing values in the data. We use BIC because it penalizes the addition of more parameters to the model, which is desirable since we have many covariates***

```{r warning=FALSE}
m0.2 <- step(m0, direction="both", 
             k = log(nrow(base)), 
             trace = FALSE)
summary(m0.2)
cat('no. coefficients m0.1:', length(coef(m0.1)),'\n')
cat('no. coefficients m0.2:', length(coef(m0.2)),'\n')
```

We can compare the two initial models with a deviance test because they are nested models.
`m0.1` model improves the `m0.2` model. We select **`m0.1`** model.

```{r warning=FALSE}
anova(m0.2, m0.1, test = 'Chisq')
```


## Validation

Standard model diagnostics using `residualPlots`

```{r warning=FALSE}
residualPlot (m0.1)
residualPlots(m0.1, layout = c(4,3), ask = FALSE, tests = FALSE)
```

***The obtained model is quite good, but we will try to improve it.***

# Improve the model: *waist*

## Aggregation

We did an aggregation using variable **`waist`**

```{r warning=FALSE}

# Agregación con múltiples variables
dfbase <- with(base, aggregate(x = cbind(ypos = Cholesterol.Category, yneg = 1 - Cholesterol.Category),
                               by = list(Insulin.Levels = Insulin.Levels, Age = Age, BMI = BMI, Blood.Pressure = Blood.Pressure),
                               FUN = sum))

```

## Simple Model

Model with variable *waist* using aggregate data:

```{r}
m_extended <- glm(cbind(ypos, yneg) ~ Age * Blood.Pressure + Age * BMI + Age * Insulin.Levels, 
                  data = dfbase, family = binomial(link = 'logit'))
summary(m_extended)
```

### Validation

We use **`residualPlots`**

```{r}
residualPlots(m_extended)
```

***The test indicates that the presence of a quadratic term is necessary***

## Model with polynomial terms

We fit the models up to order 4 and compare them using the deviance test, AIC, and BIC.

```{r}
m_extended.1 <- glm(cbind(ypos, yneg) ~ Insulin.Levels + Age + BMI + Blood.Pressure, 
                    data = dfbase, family = binomial(link = 'logit'))

m_extended.2 <- glm(cbind(ypos, yneg) ~ poly(Insulin.Levels, 2) + Age + BMI + Blood.Pressure, 
                    data = dfbase, family = binomial(link = 'logit'))

m_extended.3 <- glm(cbind(ypos, yneg) ~ poly(Insulin.Levels, 3) + Age + BMI + Blood.Pressure, 
                    data = dfbase, family = binomial(link = 'logit'))

m_extended.4 <- glm(cbind(ypos, yneg) ~ poly(Insulin.Levels, 4) + Age + BMI + Blood.Pressure, 
                    data = dfbase, family = binomial(link = 'logit'))

# Pruebas de devianza
anova(m_extended.1, m_extended.2, test = "Chisq")
anova(m_extended.2, m_extended.3, test = "Chisq")
anova(m_extended.3, m_extended.4, test = "Chisq")

# Comparación de AIC y BIC
cbind(
  AIC = AIC(m_extended.1, m_extended.2, m_extended.3, m_extended.4),
  BIC = BIC(m_extended.1, m_extended.2, m_extended.3, m_extended.4)[, 2]
)
```

***According to deviance and AIC, we would choose the 4th-order polynomial. According to BIC, the 3rd-order polynomial***

### Validation

We check 3rd and 4th order polynomial

```{r warning=FALSE}
residualPlots(m_extended.3, layout=c(1, 2))
residualPlots(m_extended.4, layout=c(1, 2))
```

## Model with 2 slopes

We fit a model with two slopes

***We are testing this model because the graph gives the impression that this pattern exists. The model provides a better BIC compared to the previous models***

```{r}
dfbase$Iwaist = as.factor(ifelse(dfbase$waist<=105,0,1))
summary(dfbase$Iwaist)

##-- Interaction with waist (two slopes?)
m_waist.5 <- glm(cbind(ypos, yneg) ~ waist * Iwaist, 
                 data = dfbase, family = binomial)  
m_waist.6 <- glm(cbind(ypos, yneg) ~ Iwaist + waist:Iwaist + poly(waist,4), 
                 data = dfbase, family = binomial)

anova(m_waist.1, m_waist.5,test='Chisq')
anova(m_waist.5, m_waist.6,test='Chisq')

cbind(AIC = AIC(m_waist.3, m_waist.4, m_waist.5, m_waist.6),
      BIC = BIC(m_waist.3, m_waist.4, m_waist.5, m_waist.6)[,2])
```

### Validation

Best diagnosis

```{r warning=FALSE}
residualPlots(m_waist.5)
```

Effect using `allEffects` to see the two slopes

```{r warning=FALSE}
plot(allEffects(m_waist.5))
```

## Full model including waist with 2 slopes

```{r}
base$Iwaist = as.factor(ifelse(base$waist<=105,0,1))

m1 <- update(m0.1, . ~ . + Iwaist + waist:Iwaist)
```

### Validation

```{r}
residualPlots(m1)
```

# Improve the model: cigars

## Aggregation

We did an aggregation using variable **`cigars`**

```{r warning=FALSE}
dfbase1 <- with(base,aggregate(cbind(ypos = Phealth, yneg = 1-Phealth),
                               list(cigars=cigars),
                               sum))
```

## Simple Model

Model with variable *cigars* using aggregate data:

```{r}
m_cigars <- glm(cbind(ypos,yneg) ~ cigars, data=dfbase1, family=binomial(link='logit'))
summary(m_cigars)
```

### Validation

We use **`residualPlots`**

```{r}
residualPlot(m_cigars)
```

## Models with polynomial terms

We fit the models up to order 4 and compare them using the deviance test, AIC, and BIC.

```{r}
m_cigars.1 <- glm(cbind(ypos,yneg) ~ cigars,        data=dfbase1,family=binomial)
m_cigars.2 <- glm(cbind(ypos,yneg) ~ poly(cigars,2),data=dfbase1,family=binomial)
m_cigars.3 <- glm(cbind(ypos,yneg) ~ poly(cigars,3),data=dfbase1,family=binomial)
m_cigars.4 <- glm(cbind(ypos,yneg) ~ poly(cigars,4),data=dfbase1,family=binomial)

# Deviance test
anova(m_cigars.1, m_cigars.2, test="Chisq")
anova(m_cigars.2, m_cigars.3, test="Chisq")
anova(m_cigars.3, m_cigars.4, test="Chisq")

# AIC and BIC
cbind(      AIC(m_cigars.1,m_cigars.2,m_cigars.3,m_cigars.4),
      BIC = BIC(m_cigars.1,m_cigars.2,m_cigars.3,m_cigars.4)[,2])
```

***According to deviance and AIC, we would choose the 4th-order polynomial. According to BIC, the 3rd-order polynomial***

### Validation

We check 3rd and 4th order polynomial

```{r warning=FALSE}
summary(m_cigars.3)
summary(m_cigars.4)
residualPlots(m_cigars.3, layout = c(1, 2))
residualPlots(m_cigars.4, layout = c(1, 2))
```

## Full model including cigars with polynomial term

```{r}
m2 <- update(m0.1, . ~ . + Iwaist + waist:Iwaist - cigars + poly(cigars,4))
```

### Validation

```{r}
residualPlots(m2)
```
