---
title: "ME - GIA. Count data (Offset)"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  html_document:
    code_folding: show
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
---


# Introduction

The ***Insurance*** data of the *MASS* package corresponds to the collection of the number of customer claims registered by an American vehicle insurer.

**Outcome**

- **_Claims_**: number of claims


**Explanatory variables**

- **_District_**: district (1:small city - 4:big city)

- **_Group_**: type of car (< 1 litre; 1-1.5; 1.5-2; >2 litres)

- **_Age_**: age (<25;25-29;30-35;>35)

- **_Holders_**: number of holders

```{r warning=FALSE, message=FALSE}
library(MASS,warn.conflicts = FALSE,quietly = TRUE) # Insurance; glm.nb
library(AER,warn.conflicts = FALSE,quietly = TRUE)  # dispersion.test

# Load data from package MASS
data(Insurance) 

df       <- data.frame(Insurance , 
                       logholders = log(Insurance$Holders))   # logholders --> offset
df$Group <- factor(df$Group,ordered = FALSE)
```


# Descriptive analysis

```{r}
summary(Insurance)
hist(Insurance$Claims, breaks = 20)
```

# Models


## Poisson Model without offset

```{r}
mod_pois_1 <- glm(Claims ~ Group, 
                    family = poisson, 
                    data   = df)
summary(mod_pois_1)
```


## Poisson Model with offset

```{r}
mod_pois_2 <- glm(Claims ~ Group + offset(logholders), 
                  family = poisson, 
                  data   = df)
summary(mod_pois_2)
```

**Overdispersion**

```{r}
dispersiontest(mod_pois_1, trafo = 1)
dispersiontest(mod_pois_2, trafo = 1)
```


